---
title: "Homework 4: EM and HMMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Homework 4: EM and HMMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2, bench}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  collapse = TRUE,
  rmarkdown.html_vignette.check_title = FALSE,
  comment = "#>"
)
```


```{r packages}
library(STATS230)
library(ggplot2)
library(bench)
```

## Problem 1: Mixture of Normals

We consider a mixture of two normal distributions with unknown means $\mu_1, \mu_2$ and known variances $\sigma_1 = 1.0$, $\sigma_2 = 0.8$. Pairs $(x_1, y_1), \ldots, (x_n, y_n)$ are i.i.d. with joint distribution

$$
\Pr(x_i, y_i) 
= [\alpha f(y_i, \mu_1, \sigma_1)]^{\mathbf{1}\{x_i = 1\}}
  [(1 - \alpha) f(y_i, \mu_2, \sigma_2)]^{\mathbf{1}\{x_i = 2\}},
$$

where $x_i \in \{1,2\}$ is a class membership variable that indicates which distribution gave rise to $y_i \in \mathbb{R}$,  

$$
\Pr(x_i = 1) = 1 - \Pr(x_i = 2) = \alpha,
$$

and

$$
f(x, \mu, \sigma) 
= \frac{1}{\sigma \sqrt{2\pi}} 
  \exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right).
$$

We also assume that we observe $y = (y_1, \ldots, y_n)$ but do not observe $x = (x_1, \ldots, x_n)$. Our goal is to infer $\alpha$, $\mu_1$, and $\mu_2$ by maximizing the observed data log-likelihood

$$
\ell(\alpha, \mu_1, \mu_2)
= \sum_{i=1}^n 
  \ln\left[ \alpha f(y_i, \mu_1, \sigma_1)
  + (1 - \alpha) f(y_i, \mu_2, \sigma_2) \right].
$$

Instead of maximizing the log-likelihood directly, we want to derive an EM algorithm to get MLEs.


### (a) Complete log-likelihood

The complete-data likelihood, taking the product over i.i.d. variables, becomes

\[
\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2)
= \prod_{i=1}^n
\Big[
\alpha f(y_i \mid \mu_1, \sigma_1)
\Big]^{\mathbf{1}\{x_i=1\}}
\Big[
(1-\alpha) f(y_i \mid \mu_2, \sigma_2)
\Big]^{\mathbf{1}\{x_i=2\}}.
\]

Taking logs gives

\[
\ln\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2)
=
\sum_{i=1}^n
\mathbf{1}\{x_i=1\}
\Big(
\ln \alpha + \ln f(y_i \mid \mu_1,\sigma_1)
\Big)
+
\mathbf{1}\{x_i=2\}
\Big(
\ln (1-\alpha) + \ln f(y_i \mid \mu_2,\sigma_2)
\Big).
\]

Using
\[
\ln f(y \mid \mu,\sigma)
=
-\ln(\sigma\sqrt{2\pi})
-\frac{(y-\mu)^2}{2\sigma^2},
\]

the complete log-likelihood becomes

\[
\begin{aligned}
\ln\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2)
&=
\sum_{i=1}^n
\mathbf{1}\{x_i=1\}
\left[
\ln\alpha
-\ln(\sigma_1\sqrt{2\pi})
-\frac{(y_i-\mu_1)^2}{2\sigma_1^2}
\right] \\
&\quad+
\sum_{i=1}^n
\mathbf{1}\{x_i=2\}
\left[
\ln(1-\alpha)
-\ln(\sigma_2\sqrt{2\pi})
-\frac{(y_i-\mu_2)^2}{2\sigma_2^2}
\right].
\end{aligned}
\]

### (b) E-step of the EM algorithm

We want to derive an expression for
\[
\mathbb{E}\!\left[
\ln\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2)
\;\middle|\;
\vec y, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}
\right].
\]

Here the expectation is taken with respect to the conditional distribution
\[
\Pr(\vec x \mid \vec y, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}).
\]
That is, we treat $\vec y$ and the current parameter values 
$\alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}$ as fixed, and average only over the
latent class indicators $\vec x$.

Because the complete log-likelihood is linear in the indicator variables
$\mathbf{1}\{x_i = 1\}$ and $\mathbf{1}\{x_i = 2\}$, taking expectations simply
replaces each indicator with its conditional probability.

Define
\[
p_i^{(k)} 
=
\Pr(x_i = 1 \mid y_i, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)})
\]

For notational convenience. Then we have
\[
\mathbb{E}[\mathbf{1}\{x_i=1\} \mid y_i] = p_i^{(k)}, 
\qquad
\mathbb{E}[\mathbf{1}\{x_i=2\} \mid y_i] = 1 - p_i^{(k)}.
\]

Therefore the expected complete log-likelihood becomes
\[
\begin{aligned}
\mathbb{E}\left[\ln\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2) \mid \vec y, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}\right]
&=
\sum_{i=1}^n
p_i^{(k)}
\left[
\ln\alpha
-\ln(\sigma_1\sqrt{2\pi})
-\frac{(y_i-\mu_1)^2}{2\sigma_1^2}
\right] \\
&\quad+
\sum_{i=1}^n
(1-p_i^{(k)})
\left[
\ln(1-\alpha)
-\ln(\sigma_2\sqrt{2\pi})
-\frac{(y_i-\mu_2)^2}{2\sigma_2^2}
\right].
\end{aligned}
\]

The important idea is that the only randomness in the complete-data log-likelihood comes from the unobserved labels $x_i$.
The idea behind EM here is to replace unknown class labels with their current probabilities, then maximize a weighted log-likelihood to get new parameter estimates, and repeat until convergence.

## (c) M-step of the EM algorithm

To maximize, we take partial derivatives with respect to $\alpha$, $\mu_1$, and $\mu_2$.

With respect to $\alpha$, we have
$$
\frac{\partial}{\partial \alpha}
\mathbb{E}\left[\ln\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2) \mid \vec y, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}\right]
=
\sum_{i=1}^n
\frac{p_i^{(k)}}{\alpha}
-
\sum_{i=1}^n
\frac{1-p_i^{(k)}}{1-\alpha}
=
0.
$$

Solving for $\alpha$, we get
$$
\alpha^{(k+1)} = 
\frac{\sum_{i=1}^n p_i^{(k)}}{n}.
$$

Checking the second derivative, we have
$$
\frac{\partial^2}{\partial \alpha^2}
=
-\sum_{i=1}^n
\frac{p_i^{(k)}}{\alpha^2}
-
\sum_{i=1}^n
\frac{1-p_i^{(k)}}{(1-\alpha)^2}
<
0,
$$
so the critical point is a maximum.

With respect to $\mu_1$, we have
$$
\frac{\partial}{\partial \mu_1}
\mathbb{E}\left[\ln\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2) \mid \vec y, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}\right]
=
-\sum_{i=1}^n
p_i^{(k)}
\frac{(y_i - \mu_1)}{\sigma_1^2}
=
0.
$$

Solving for $\mu_1$, we get
$$
\mu_1^{(k+1)} = 
\frac{\sum_{i=1}^n p_i^{(k)} y_i}{\sum_{i=1}^n p_i^{(k)}}.
$$

Checking the second derivative, we have
$$
\frac{\partial^2}{\partial \mu_1^2}
=
-\sum_{i=1}^n
\frac{p_i^{(k)}}{\sigma_1^2}
<
0,
$$
so the critical point is a maximum.

With respect to $\mu_2$, we have
$$
\frac{\partial}{\partial \mu_2}
\mathbb{E}\left[\ln\Pr(\vec x, \vec y \mid \alpha, \mu_1, \mu_2) \mid \vec y, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}\right]
=
-\sum_{i=1}^n
(1-p_i^{(k)})
\frac{(y_i - \mu_2)}{\sigma_2^2}
=
0.
$$

Solving for $\mu_2$, we get
$$
\mu_2^{(k+1)} = 
\frac{\sum_{i=1}^n (1-p_i^{(k)}) y_i}{\sum_{i=1}^n (1-p_i^{(k)})}.
$$

Checking the second derivative, we have
$$
\frac{\partial^2}{\partial \mu_2^2}
=
-\sum_{i=1}^n
\frac{1-p_i^{(k)}}{\sigma_2^2}
<0,
$$
so the critical point is a maximum.

Therefore, the EM algorithm is given by the following updates:
\[
\begin{aligned}
p_i^{(k)}
&=
\Pr(x_i = 1 \mid y_i, \alpha^{(k)}, \mu_1^{(k)}, \mu_2^{(k)}) \\
&=
\frac{\alpha^{(k)} f(y_i \mid \mu_1^{(k)}, \sigma_1)}{\alpha^{(k)} f(y_i \mid \mu_1^{(k)}, \sigma_1) + (1-\alpha^{(k))} f(y_i \mid \mu_2^{(k)}, \sigma_2)}, \
\qquad
i = 1, \ldots, n, \\
\alpha^{(k+1)}
&=
\frac{\sum_{i=1}^n p_i^{(k)}}{n}, \\
\mu_1^{(k+1)}
&=
\frac{\sum_{i=1}^n p_i^{(k)} y_i}{\sum_{i=1}^n p_i^{(k)}}, \\
\mu_2^{(k+1)}
&=
\frac{\sum_{i=1}^n (1-p_i^{(k)}) y_i}{\sum_{i=1}^n (1-p_i^{(k)})}.
\end{aligned}
\]

## (d) Implementation and example using `mixture_data.txt`

The EM algorithm is implemented in the function `em_mixture_normals()`. The function takes as input a vector of observations `y`, initial parameter values `alpha_init`, `mu1_init`, and `mu2_init`, and a maximum number of iterations `max_iter`. The function returns a list containing the final parameter estimates and the log-likelihood at each iteration.

```{r example}
y <- as.vector(read.table(system.file("mixture_data.txt",
                                      package = "STATS230")))[[1]]
rslt <- em_mixture_normals(y)
print(rslt)
```

## (e) Simulation study

Now, we will test the performance of our EM algorithm by simulating data from a known mixture of normals and checking if we can recover the true parameters.

```{r simulation}
set.seed(123)

# True parameters
alpha_true <- 0.3
mu1_true <- 2.1
mu2_true <- 5.2
sigma1 <- 1
sigma2 <- 0.8
datasets <- 100
n_obs <- 300

results_list <- vector("list", length = datasets)
for (d in 1:datasets) {
  # Simulate data
  x <- rbinom(n_obs, size = 1, prob = 1 - alpha_true) + 1  # x_i = 1 or 2
  y <- ifelse(x == 1,
              rnorm(n_obs, mean = mu1_true, sd = sigma1),
              rnorm(n_obs, mean = mu2_true, sd = sigma2))

  # Run EM algorithm on simulated data
  rslt_sim <- em_mixture_normals(y,
                                 mu1_init = quantile(y, 0.25),
                                 mu2_init = quantile(y, 0.75),
                                 sigma1 = sigma1, sigma2 = sigma2)

  # Store results
  results_list[[d]] <- list(
    alpha = rslt_sim$alpha,
    mu1 = rslt_sim$mu1,
    mu2 = rslt_sim$mu2
  )
}
```


Now, we plot the results from above:
```{r plot, echo = FALSE, message = FALSE, warning = FALSE}
# Put results into data frame
results_df <- do.call(rbind, lapply(results_list, as.data.frame))

# Density plots of the parameter estimates across simulations
library(ggplot2)
results_long <- reshape2::melt(results_df, variable.name = "Parameter",
                               value.name = "Value")

# Create data frame for vertical lines
vline_data <- data.frame(
  Parameter = c("alpha", "mu1", "mu2"),
  xint = c(alpha_true, mu1_true, mu2_true),
  label = c("True α", "True μ₁", "True μ₂")
)

# Create data frame to set x-axis limits for each panel
bounds_data <- data.frame(
  Parameter = c("alpha", "alpha", "mu1", "mu1", "mu2", "mu2"),
  Value = c(0, 1, 1, 6, 1, 6)
)

# Create labeller for facets with expressions
param_labels <- c(
  alpha = "alpha",
  mu1 = "mu[1]",
  mu2 = "mu[2]"
)

# Plot
ggplot(results_long, aes(x = Value, fill = Parameter)) +
  geom_blank(data = bounds_data) +
  geom_histogram(aes(y = ..density..), position = "identity",
                 alpha = 0.5, bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~ Parameter, scales = "free", 
             labeller = as_labeller(param_labels, label_parsed)) +
  geom_vline(data = vline_data, aes(xintercept = xint), 
             linetype = "dashed", color = "red") +
  scale_fill_discrete(labels = c(expression(alpha), expression(mu[1]), expression(mu[2]))) +
  labs(title = "Histograms of Parameters", x = "Value", y = "Density") +
  theme_minimal()
```

Notice that, above, we initialized the EM algorithm with the 25th and 75th percentiles of the data for `mu1_init` and `mu2_init`, respectively.
Let's see how the results change if we do the opposite:

```{r sim2}
# Run EM algorithm with reversed initializations
results_list_reversed <- vector("list", length = datasets)
for (d in 1:datasets) {
  # Simulate data
  x <- rbinom(n_obs, size = 1, prob = 1 - alpha_true) + 1  # x_i = 1 or 2
  y <- ifelse(x == 1,
              rnorm(n_obs, mean = mu1_true, sd = sigma1),
              rnorm(n_obs, mean = mu2_true, sd = sigma2))
  # Run EM algorithm on simulated data with reversed initializations
  rslt_sim <- em_mixture_normals(y,
                                 mu1_init = quantile(y, 0.75),
                                 mu2_init = quantile(y, 0.25),
                                 sigma1 = sigma1, sigma2 = sigma2)
  # Store results
  results_list_reversed[[d]] <- list(
    alpha = rslt_sim$alpha,
    mu1 = rslt_sim$mu1,
    mu2 = rslt_sim$mu2
  )
}
```

```{r plot2, echo = FALSE, message = FALSE, warning = FALSE}
# Put reversed results into data frame
results_df_reversed <- do.call(rbind, lapply(results_list_reversed, as.data.frame)) 
results_long_reversed <- reshape2::melt(results_df_reversed, variable.name = "Parameter",
                                        value.name = "Value")
# Plot reversed results
ggplot(results_long_reversed, aes(x = Value, fill = Parameter)) +
  geom_blank(data = bounds_data) +
  geom_histogram(aes(y = ..density..), position = "identity",
                 alpha = 0.5, bins = 30) +
  geom_density(alpha = 0.2) +
  facet_wrap(~ Parameter, scales = "free",
             labeller = as_labeller(param_labels, label_parsed)) +
  geom_vline(data = vline_data, aes(xintercept = xint),
             linetype = "dashed", color = "red") +
  scale_fill_discrete(labels = c(expression(alpha), expression(mu[1]), expression(mu[2]))) +
  labs(title = "Histograms of Parameters with Reversed Initializations", x = "Value", y = "Density") +
  theme_minimal()
```

Now, everything's reversed! This is known as the "label switching" problem because the likelihood is symmetric in $\mu_1$ and $\mu_2$, so the EM algorithm can converge to either mode depending on the initialization. To fix this issue, we can enforce a consistent labeling convention, such as always labeling the component with the smaller mean as component 1. This way, we will always have $\mu_1 < \mu_2$, and the algorithm will converge to the same mode regardless of initialization.

Going back to our original results, the question asked for the bias of $\hat \alpha$ and its associated Monte Carlo error:
```{r bias, echo = FALSE}
# Compute bias and Monte Carlo error for alpha
alpha_estimates <- sapply(results_list, function(x) x$alpha)
alpha_bias <- mean(alpha_estimates) - alpha_true
alpha_mc_error <- sd(alpha_estimates) / sqrt(datasets)
cat("Bias of alpha:", alpha_bias, "\n")
cat("Monte Carlo error of alpha:", alpha_mc_error, "\n")
```

The bias is small relative to $\alpha$, and the Monte Carlo error is larger than the bias, suggesting that the estimator is unbiased.

# Problem 2: Hidden Markov Model (Occasionally Dishonest Casino)

The setup of this question is that we are at a casino that sometimes uses a fair die, and sometimes uses a loaded die, and they randomly switch between the two.
Initially, the casino selects one of the dice uniformly randomly (i.e., with 1/2 probability for each).
We are given the transition probabilities between the two states (1 = fair die, 2 = loaded die) and the probabilities of observing each outcome (1-6) given the state.

## (a) Simulating the Casino

The function `occasionally_dishonest_casino()` simulates a sequence of die rolls from the casino. It takes as input the number of rolls to simulate and returns a list containing the sequence of states (which die was used) and the sequence of observed outcomes.

```{r casino}
set.seed(123)

hmm_sim <- occasionally_dishonest_casino(n_obs = 100)

# plot the sequence of states and outcomes
state_outcome_df <- data.frame(
  Time = seq_along(hmm_sim$states),
  State = factor(hmm_sim$states, levels = 1:2, labels = c("Fair", "Loaded")),
  Outcome = hmm_sim$observations
)

ggplot(state_outcome_df, aes(x = Time)) +
  geom_line(aes(y = as.numeric(State)), color = "blue") +
  geom_line(aes(y = Outcome), color = "red", alpha = 0.5) +
  scale_y_continuous(
    name = "State (1=Fair, 2=Loaded)",
    sec.axis = sec_axis(~ ., name = "Outcome (1-6)")
  ) +
  labs(title = "Simulated Sequence of States and Outcomes") +
  theme_minimal()
```

## (b) Implement Forward and Backward Algorithms

The forward algorithm computes the probability of observing the sequence of outcomes up to time $t$ and being in state $j$ at time $t$. The backward algorithm computes the probability of observing the sequence of outcomes from time $t+1$ to the end given that we are in state $j$ at time $t$.
Together, these algorithms allow us to compute the probabilities of being in each state at each time point given the entire sequence of observations.
The function `forward_backward_casino()` implements these algorithms. It takes as input the sequence of observations, the initial distribution over states, the transition probabilities, and the emission probabilities, and returns a list containing the forward probabilities (alpha), backward probabilities (beta), and the transition and emission matrices.
This function is called internally by `occasionally_dishonest_casino()` to compute the forward and backward probabilities for the simulated data.

Below, we plot the true state, overlaid with the probability of being in the loaded state at each time point.
```{r forward_backward}
state_prob_df <- data.frame(
  Time = seq_along(hmm_sim$states),
  TrueState = hmm_sim$states - 1,
  ProbLoaded = hmm_sim$gamma[, 2]  # Probability of being in the loaded state
)

ggplot(state_prob_df, aes(x = Time)) +
  geom_line(aes(y = ProbLoaded), color = "red") +
  geom_line(aes(y = as.numeric(TrueState)), color = "blue") +
  scale_y_continuous(
    name = "Probability of Loaded State",
    sec.axis = sec_axis(~ ., name = "True State is Loaded (Indicator)")
  ) +
  labs(title = "Probability of Loaded State Over Time") +
  theme_minimal()
```

