---
title: "Homework 5: Monte Carlo and Markov Chains"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Homework 5: Monte Carlo and Markov Chains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  collapse = TRUE,
  rmarkdown.html_vignette.check_title = FALSE,
  comment = "#>"
)
```


```{r packages}
library(STATS230)
library(ggplot2)
```


## Monte Carlo Estimation

When the distribution of interest is complicated, we can use Monte Carlo methods to estimate moments. The `montecarlo_moments()` function estimates the $k$-th moment of a distribution using either naive Monte Carlo or importance sampling.
Naive Monte Carlo estimation involves drawing samples directly from the target distribution and computing the sample average of $X^k$.
Importance sampling, on the other hand, involves drawing samples from a proposal distribution and weighting them by the ratio of the target density to the proposal density. The `montecarlo_moments()` function allows you to specify both the target distribution and the proposal distribution, as well as the number of samples to draw.

Below, we compare the estimates of the second moment of a Gamma distribution using both methods. The target distribution is a Gamma distribution with shape 4 and rate 2, so the second moment is $E[X^2] = \frac{\alpha(\alpha + 1)}{\beta^2} = \frac{4 \cdot 5}{2^2} = 5$.
Using naive Monte Carlo:
```{r naive}
montecarlo_moments(type = "naive", moment = 2)
```

we obtain an estimate close to 5, the true second moment.

For importance sampling, one thing we should consider is the choice of proposal distribution.
Though the homework asks us to use log-normal, we need to choose the parameters of the log-normal distribution carefully to ideally give it heavier tails than the target distribution.
The success of the importance sampling estimator depends on the instrument distribution having at least a very similar support to the target distribution, and ideally heavier tails, to decrease variance of the estimator.

Here, we can try to match the moments of the target distribution to those of the proposal, which is a fairly standard approach..
The mean of the target distribution is $\frac{\alpha}{\beta} = \frac{4}{2} = 2$, and the variance is $\frac{\alpha}{\beta^2} = \frac{4}{4} = 1$.
We can try to choose parameters for the log-normal distribution such that the mean is around 2 and the variance is around 1, or slightly larger to give it heavier tails.
Here, we can set the meanlog to $\log(2) - \log(1.25)^2 / 2$ and the sdlog to $\log(1.25)$.

Comparing the two densities,
```{r densities}
x <- seq(0, 10, length.out = 1000)

dens_df <- data.frame(
  x = rep(x, 2),
  density = c(
    dgamma(x, shape = 4, rate = 2),
    dlnorm(x, meanlog = log(2) - log(1.25)^2 / 2, sdlog = log(1.25))
  ),
  dist = factor(
    rep(c("Target", "Proposal"), each = length(x)),
    levels = c("Target", "Proposal")
  )
)

ggplot(dens_df, aes(x = x, y = density, color = dist)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Target" = "blue", "Proposal" = "red")) +
  labs(x = "x", y = "Density", color = NULL) +
  theme_minimal()
```

We can see that this might not be a great choice for the proposal distribution. Let's test it, anyway.
Using importance sampling with this proposal distribution:

```{r importance}
montecarlo_moments(type = "importance", moment = 2)
```

Notice that the Monte Carlo error for the importance sampling estimator is much larger than that of the naive Monte Carlo estimator.
This is not inherently a property of importance sampling, rather it's about the choice of proposal distribution.
As we saw above, the proposal distribution has areas of extremely low density where the target distribution has non-negligible density, which leads to very large importance weights and thus high variance in the estimator.
We can mess with the parameters a little bit to try to get a better proposal:

```{r densities2}
x <- seq(0, 10, length.out = 1000)

dens_df <- data.frame(
  x = rep(x, 2),
  density = c(
    dgamma(x, shape = 4, rate = 2),
    dlnorm(x, meanlog = log(4) - 1^2 / 2, sdlog = 1)
  ),
  dist = factor(
    rep(c("Target", "Proposal"), each = length(x)),
    levels = c("Target", "Proposal")
  )
)

ggplot(dens_df, aes(x = x, y = density, color = dist)) +
  geom_line(linewidth = 1.1) +
  scale_color_manual(values = c("Target" = "blue", "Proposal" = "red")) +
  labs(x = "x", y = "Density", color = NULL) +
  theme_minimal()
```

And trying again with these new parameters,

```{r importance2}
montecarlo_moments(
  type = "importance",
  moment = 2,
  instrument_distribution = list(sampler = function(n) rlnorm(n, meanlog = log(4) - 1^2 / 2, sdlog = 1),
                                 density = function(x) dlnorm(x, meanlog = log(4) - 1^2 / 2, sdlog = 1))
)
```

we get an estimator with smaller variance than the naive Monte Carlo estimator.

## Wright-Fisher Markov Chain

We want to show
\[
\mu \equiv \sum_{i=0}^{2m} i\pi_i = \frac{2mu}{u+v}.
\]
Let \(x = (0,1,2,\dots,2m)^T\). Since \(\pi\) is stationary,
\[
\mu = \pi^T x = \pi^T P x,
\]
where \(P\) is the transition matrix.

For state \(i\),
\[
(Px)_i = \sum_{j=0}^{2m} j\,P_{ij}.
\]
Here,
\[
P_{ij}
= \binom{2m}{j}
\left(\frac{i(1-v)+(2m-i)u}{2m}\right)^j
\left(1-\frac{i(1-v)+(2m-i)u}{2m}\right)^{2m-j},
\]
so \(P_{ij}\) is Binomial\((2m,p_i)\) with
\[
p_i=\frac{i(1-v)+(2m-i)u}{2m}.
\]
Therefore,
\[
(Px)_i = \mathbb{E}[J\mid i] = 2m\,p_i = i(1-v)+(2m-i)u.
\]

Now compute:
\[
\mu
= \pi^T P x
= \sum_{i=0}^{2m}\pi_i (Px)_i
= \sum_{i=0}^{2m}\pi_i\big(i(1-v)+(2m-i)u\big).
\]
So
\[
\mu
= (1-v)\sum_{i=0}^{2m} i\pi_i
  + u\sum_{i=0}^{2m}(2m-i)\pi_i
= (1-v)\mu + u(2m-\mu).
\]
Hence
\[
\mu = \mu + 2mu - (u+v)\mu
\quad\Longrightarrow\quad
(u+v)\mu = 2mu
\quad\Longrightarrow\quad
\mu = \frac{2mu}{u+v}.
\]

Next, we confirm this numerically using the function `wright_fisher_path()`, which simulates a path of the Wright-Fisher Markov chain and returns the states visited.
We can simulate a long path and compute the empirical mean of the states visited, which should be close to the theoretical mean we derived above.
If we set \(m=50\), \(u=0.35\), and \(v=0.5\), then the theoretical mean is \(\frac{2 \cdot 50 \cdot 0.35}{0.35 + 0.5} = \frac{35}{0.85} \approx 41.18\).
```{r wright-fisher}
set.seed(123)
m <- 50
u <- 0.35
v <- 0.5
N <- 2 * m

path <- wright_fisher_path(n_steps = 10000, x0 = 0, N = N, u = u, v = v)
plot(path, type = "l", xlab = "Step", ylab = "State")

theoretical_mean <- N * u / (u + v)
full_mean <- mean(path)
burn_in <- 2000
post_burn_in_mean <- mean(path[(burn_in + 1):length(path)])

c(
  theoretical = theoretical_mean,
  empirical_full = full_mean,
  empirical_post_burn_in = post_burn_in_mean
)
```